---
title: 'Chapter 5: Surprises'
author: "Andrew Flowers"
date: "3/7/2017"
output: html_document
---

Software developers don't like to be surprised. So no wonder many are frustrated with R. It behaves in ways inconsistent with more traditional programming languages. This behavior can sometimes be chalked up to the idiosyncratic designs within R. But, frankly, some of its behavior is inconsistent or outright confusing.

### Function masking

_\"There are only two hard things in Computer Science: cache invalidation and naming things. -- Phil Karlton_

Yes, that's a programming joke. (Hilarious, right? OK, I'll stop.) The joke has a lot of truth to it: naming things is hard.

In R, you'll probably run into problems with your namespace. In particular, you'll load two identially-named functions from different pacakges. And it will call all sorts of problems.

A classic instance of this comes from using multiple generations of Hadley Wickham's data manipulation packages. Using both `plyr` and `dplyr`, in particular, illustrates the problem of "function masking" -- overwriting the use of one function with another one with the same name. Pay attention to the warning messages.

```{r cache = TRUE}
library(dplyr)
library(plyr)
```

Now see what happens when use the `mtcars` dataset and perform a simple calculation using the `summarize()` function.

```{r message = FALSE}
library(dplyr)
library(plyr)
# summarize() is "masked""
mtcars %>% 
  group_by(cyl) %>% 
  summarize(avg_mpg = mean(mpg))
```

Now that is **not** what we wanted! It was supposed to return the `avg_mpg` for cars grouped by the number of cylinders (`cyl`). 

To specify which `summarize()` function you want to use exactly, preface your function with package name and two colons (`:`), like this:

```{r}
# summarize() from dplyr
mtcars %>% 
  group_by(cyl) %>% 
  dplyr::summarize(avg_mpg = mean(mpg))
```
### Annoying levels

Note: This "mistake" suggestion comes from Jose Manuel Vera. 

Factors are an R data type that often causes problems (as we covered in chapter 2). As you'll see in the example below, we'll create a dummy dataframe that has a `letters` column that are factors.

```{r cache = TRUE}
df <- data.frame(letters=letters[1:5],numbers=seq(1:5))

df

str(df)
```

With factors, you can discover the pre-specified categories with the `levels()` function. But it's a mistake to assume these levels disappear once said dataframe is subsetted.

```{r cache = TRUE}
levels(df$letters)
subdf <- subset(df, numbers <=3)
subdf
# check levels
levels(subdf$letters)
```

Weird -- you'd think that the levels "d" and "e" would disappeard after subetting out those vaules. But they don't. To permanently remove unneeded levels use the `droplevels()` function.

```{r}
droplevels(subdf$letters)
```

### Escape characters

Like any programming language, R supports regular expressions -- a special syntax and set of methods for searching strings. You may use the `regexpr()` function, or one of its siblings like `gsub()` or `grep()`. But in using these functions, you'll have to use an escape character -- a character that tells the function to interpret the _subsequent_ character differently.  In R, as in most languages, the escape character is the backslash (`\`).

Take this examle: we're trying to remove the `$` characters from a vector of prices. (So that the resulting data is just numeric data.) We'll do this using the `gsub()` function -- removing every instance of `$`, or replace it with the empty character `""`. Type `?gsub` to learn more about this function.

```{r cache = TRUE}
prices <- c("$217", "$840", "$503")

# Incorrect: No escape character \
gsub("$", "", prices)
```
Ah, well that did precisely nothing. Let's add the escape character before the `$`.

```{r error = TRUE}
# Incorrect: Only one escape character \
gsub("\$", "", prices)
```
Scratching your head? That's because in R the backslash _itself_ needs to be escaped. 

So often the correct number of escape characters is two `\\`. 
```{r}
# Correct: two escape characters \\
gsub("\\$", "", prices)
```
### Lists and double-bracket notation `[[]]`

Of R's data types, a list is the most flexible. It can take hetergenous input (a numeric, a charactacter and a boolean all at once). But accessing values within a list often comes as a surprise for beginners.

```{r cache = TRUE}
customers <- list(
  names = c("Jen", "Mark", "Mindy"), 
  age = c(41, 17, 33), 
  female = c(TRUE, FALSE, TRUE)
  )

# Extract ages
customers[2]

# But it's a list!?
str(customers[2])
```

When subsetting a list using single-bracket notation (`[ ]`), another list is returned. To get the actual atomic data type held with that list you need to use double-bracket notation (`[[ ]]`).

```{r}
# Extract ages as a numeric vector
customers[[2]]

str(customers[[2]])
```